---
title: "Introduction to tablestylizer"
author: "Ville Langén"
date: "`r Sys.Date()`"
#output: rmarkdown::html_vignette
output:
  rmarkdown::html_vignette:
    fig_caption: false
    toc: true
    toc_depth: 1
    df_print: default

vignette: >
  %\VignetteIndexEntry{Introduction to tablestylizer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

```

```{r setup, echo=FALSE, message=FALSE}
library(tablestylizer)
library(tableone)
library(tibble)
library(dplyr)
library(flextable)
library(knitr)
```


<br/><br/>

# 1. Overview {#overview}

**tablestylizer** provides a set of functions designed to enhance the styling and formatting of tables in R. This package is particularly useful for making tables suitable for publication or presentation. The main functionalities of **tablestylizer** can be summarized through the following four functions:

- `stylize_tableone()`: This function processes a tableone object (that has been generated by the CreateTableOne function of the [tableone](https://cran.r-project.org/web/packages/tableone/vignettes/introduction.html) package) and performs various operations to make the output more suitable for publication.

- `stylize_p()`: Rounds and formats numeric values for improved readability, designed for presenting p-values in statistical outputs.

- `stylize1()`: Stylizes numeric values with highly custom rules aiming for 1 decimal place.

- `stylize2()`: Stylizes numeric values with highly custom rules aiming for 2 decimal places.

<br/>
<br/>

# 2. Quick showcase: from raw data to publication-ready table

This section demonstrates the unparalleled ease and efficiency of creating publication-ready tables using the **tableone**, **flextable**, and **tablestylizer** packages. With just a few simple commands, users can transform raw data into visually appealing and professionally formatted tables suitable for academic journals and similar purposes.


```{r, ft.align="left"}
# Load the example dataset
data("tablestylizer_df", package = "tablestylizer")

## Vectors for 1) all, 2) categorical variables 
myVars <- c("Age (years)", "Women", "Height (inches)", "Likes coffee", "Favorite color")
catVars <- c("Women", "Likes coffee", "Favorite color")

## Create and store a tableone object
my_df <- CreateTableOne(vars = myVars, data = tablestylizer_df, factorVars = catVars)

# Process the tableone object with stylize_tableone()
my_df <- stylize_tableone(my_df)

# Create a flextable object
ft <- my_df %>% flextable() %>% width(width = 2.5) %>%
                  align(j = 2, align = "center", part = "all") %>% 
                  bold(part = "header", bold = TRUE) %>% 
                  add_header_lines("Table 1. Characteristics of the (imaginary) study participants.") %>% 
                  add_footer_lines(("Values are means (and standard deviations) for continuous data and numbers (and percentages) for categorical data.")) %>%
                  prepend_chunks(i = 7:9, j = 1, as_chunk("\t"))

# Display the ready table
ft                                
```


<br/>
<br/>


# 3. Motivations and limitations

## 3.1 Motivation

The motivations for developing the **tablestylizer** package include, but are not limited to, the following:

1. **Enhancing Tableone Functionality**: While the [tableone](https://cran.r-project.org/web/packages/tableone/vignettes/introduction.html) package is a valuable tool for generating nearly publication-ready tables of characteristics, additional tidying up is often required before submission to a journal.

2. **Versatility Across Table Types**: Many functions in the **tablestylizer** package are not limited to "Characteristics" tables but can be applied to any tables that require tasks such as rounding, offering a comprehensive set of tools for users working with tabular data in R.

3. **Eliminating Manual Tweaking**: The **tablestylizer** package aims to nullify the need for transferring data to external spreadsheet programs for manual tweaking, streamlining the data formatting process within R itself.

## 3.2 Limitations

Despite its versatility, the **tablestylizer** package comes with certain limitations:

1. **Rounding Methodology:** The rounding method used in this package follows deliberately the traditional arithmetic rounding, where numbers ending in .5 always round upwards. I call this "high school rounding" reminiscent of the rounding method most of us learn in school. This approach, similar to what many popular programs like Excel employ, contrasts with Base R, which uses [banker's rounding](https://wiki.c2.com/?BankersRounding).

2. **stylize_tableone() Function:** At present, the `stylize_tableone()` function is designed to produce means ± SDs and numbers (percentages) statistics. It does not provide medians (IQRs). If the capability to generate medians (IQRs) is a desired feature, it may be considered for implementation in future updates of the package.


<br/><br/>


# 4. Example datasets for practicing

To facilitate the exploration and application of the `stylize_tableone()` function, the **tablestylizer** package includes several example datasets.

<br/>

Here's a breakdown of all example datasets in the **tablestylizer** package:

| Dataset name     | Format               | Description                                                                    |
| :--------------- | :------------------- | :----------------------------------------------------------------------------- |
| tablestylizer_df| tibble                | Example Dataset with Imaginary Demographic Data|
| tableone_1g     | tableone object (list)| Example tableone Object where data are presented in only one column ("Overall")|
| tableone_3g     |tableone object (list) | Example tableone Object where data are presented separately for subgroups|
| rtableone_3g_p  | tableone object (list)| Same as the tableone_3g, but including    between-group comparison test|

<br/>

Let's skim through the first example dataset.

```{r, echo=TRUE}
# Load the example dataset
data("tablestylizer_df", package = "tablestylizer")

# Display the head of the dataset
head(tablestylizer_df)
```

<br/>

The data presented above can easily be converted into a tableone object by running a code that uses the function `CreateTableOne` from the **tableone** package.

```{r, echo=TRUE}
## Get variables names
dput(names(tablestylizer_df))

## Vector of all variables that shall be summarized by CreateTableOne
myVars <- c("Age (years)", "Women", "Height (inches)", "Likes coffee", "Favorite color", "Group")

## Vector of categorical variables
catVars <- c("Women", "Likes coffee", "Favorite color", "Group")

## Create and display tableone object
CreateTableOne(vars = myVars, data = tablestylizer_df, factorVars = catVars)
```

<br/>

For the sake of further practice sessions, the above tableone object is available as an example dataset object called `tableone_1g` in the **tablestylizer** package.



<br/><br/>


# 5. Tidying up tableone objects: `stylize_tableone()`

The `stylize_tableone()` function in the **tablestylizer** package is designed to enhance the presentation of tableone objects, making them more suitable for publication or presentation.

Here's a list of what `stylize_tableone()` does:

- **Moving Percentages:** The [tableone](https://cran.r-project.org/web/packages/tableone/vignettes/introduction.html) package includes annotations such as "(%)" for percentages in the variable column. However, these heavy annotations are not commonly found in published manuscripts. `stylize_tableone()` addresses this by moving percentage annotations rightwards to the corresponding statistics columns.

- **Handling Means and SDs:** Annotations like "(mean (SD))" for means with standard deviations are also repositioned rightwards and formatted as "𝑥 ± 𝑦". This modification aligns with the typical presentation style in many published works, improving the overall readability of the table.

- **P-value Formatting:** The default behavior of the `tableone` package expresses p-values with 3 digits. However, in many published works, p-values are commonly presented with 2 digits for brevity and clarity. The `stylize_tableone()` function aligns with this convention, expressing p-values with 2 digits whenever possible, providing a more standardized and visually appealing format.

4. **Changing "Overall" to "Value"**: For one group data, the text "Overall" is changed to "Value" to enhance clarity and consistency in the table presentation.


## 5.1 One column of data

Let's look at the example dataset `tableone_1g` of the **tablestylizer** package. `tableone_1g` is already in tableone object format and thus ready to be modified by the function `stylize_tableone()`.

```{r, echo=TRUE}
# Load the example dataset
data("tableone_1g", package = "tablestylizer")

# Display "tableone_1g" object
tableone_1g
```

<br/>

This tableone object can now be processed by `stylize_tableone()`.

```{r, echo=TRUE}
# Process the tableone object, pipe it to print() command to align left
stylize_tableone(tableone_1g) %>% print(right = FALSE)

```

<br/>

## 5.1.1 Recoding of binary variables

If the levels of the binary categorical factor variable need to be reversed, it can be accomplished easily. While Base R can be used for this purpose, the simplest solution is to leverage the **forcats** package.

As an example, let’s change `Likes coffee` to `Dislikes coffee`. There are numerous methods to achieve this, but one straightforward solution is to reverse the order of the factor levels.

Let’s begin by creating a new factor variable, `Likes coffee 2`, with the order of the factor levels reversed.


```{r, echo=TRUE}
# Load forcats
library(forcats)

# Load the example dataset
data("tablestylizer_df", package = "tablestylizer")

# Create a new factor variable with reversed levels
tablestylizer_df$`Likes coffee 2` <- fct_rev(tablestylizer_df$`Likes coffee`)

# Display head
head(tablestylizer_df[,c(1,5,8)], n=5)

```

<br/>

When you examine the data frame above, it might appear that nothing has changed. However, upon creating a new tableone object with that data, you'll notice that the order of the levels has been altered.

```{r}

## Vector of all variables that shall be summarized by CreateTableOne
myVars <- c("Likes coffee", "Likes coffee 2")

## Vector of categorical variables
catVars <- c("Likes coffee", "Likes coffee 2")

## Create and display tableone object
CreateTableOne(vars = myVars, data = tablestylizer_df, factorVars = catVars)

```

<br/>

Now, we know how to effortlessly reproduce the same formatted table as above, but with the `Likes coffee` variable reversed.


```{r}

# Load the example dataset
data("tablestylizer_df", package = "tablestylizer")

# Create a new factor variable with reversed levels
tablestylizer_df$`Likes coffee 2` <- fct_rev(tablestylizer_df$`Likes coffee`)

## Vector of all variables that shall be summarized by CreateTableOne
myVars <- c("Age (years)", "Women", "Height (inches)", "Likes coffee 2", "Favorite color", "Group")

## Vector of categorical variables
catVars <- c("Women", "Likes coffee 2", "Favorite color", "Group")

## Create and store a tableone object
my_df <- CreateTableOne(vars = myVars, data = tablestylizer_df, factorVars = catVars)

# Display the tableone object
my_df

# Process the tableone object
my_df <- stylize_tableone(my_df)

# Rename binary variables as desired
my_df[5,1] <- "Dislikes coffee"

# Display the processed tableone object, pipe it to print() command to align left
my_df %>% print(right = FALSE)

```


<br/>

## 5.2 Several columns of data

Let's take a closer look at the example dataset `tableone_3g` provided by the **tablestylizer** package. This dataset comprises subgroup data, presenting information across multiple columns.


```{r, echo=TRUE}
# Load the example dataset
data("tableone_3g", package = "tablestylizer")

# Display "tableone_3g" object
tableone_3g
```

<br/>

The `stylize_tableone()` function adeptly handles this type of tableone object.

```{r, echo=TRUE}
# Process the tableone object, pipe it to print() command to align left
stylize_tableone(tableone_3g) %>% print(right = FALSE)

```

<br/>


## 5.2 Between-group comparison

Now, let's examine the example dataset `tableone_3g_p` provided by the **tablestylizer** package. Similar to `tableone_3g`, this dataset contains subgroup data. However, `tableone_3g_p` includes additional between-group comparisons with p-values. The `stylize_tableone()` function seamlessly handles this type of tableone object. It's noteworthy that `stylize_tableone()` processes the p-values using a dedicated function called `stylize_p()`, which is also included in the tablestylizer package.


```{r, echo=TRUE}
# Load the example dataset
data("tableone_3g_p", package = "tablestylizer")

# Display "tableone_3g_" object
tableone_3g_p
```

<br/>

The tableone object is now ready to undergo processing by `stylize_tableone()`.

```{r, echo=TRUE}
# Process the tableone object, pipe it to print() command to align left
stylize_tableone(tableone_3g_p) %>% print(right = FALSE)

```

<br/><br/>


# 6. Demonstration of the `stylize1()` function

In this chapter, we showcase the usage of the `stylize1` function with randomly generated data.

Let's examine the following random data.

```{r, echo=FALSE, message=FALSE}

random_df <- 

tribble(
  ~Variable,      ~`Group A`,   ~`Group B`,    ~`Group C`,
  "Biomarker 1",  -0.6,         -0.2,          1.6,
  "Biomarker 2",   0.0001,       0.0003,       0.003,
  "Biomarker 3",   355141.3,     145266.8,   -794150.6,
  "Biomarker 4",   1300000,      -170000000,      170000000
)

# Output the stylized dataframe using kable
kable(random_df, caption = "random_df")

```

<br/>
<br/>


The preceding data could be presented in a more polished, publication-ready format. We will now utilize the `stylize1` function to enhance the presentation of each column's data and display the result below.

<br/>

<blockquote style="margin-left: 40px; border-left: 4px solid #ccc; padding-left: 10px;">
  Please note that the mutate_at() function from the dplyr package is the most efficient method for applying these transformations across an entire dataset.
</blockquote>

<br/>
<br/>

```{r, echo=TRUE}
stylized_data <- random_df %>%
  mutate_at(vars("Group A", "Group B", "Group C"), stylize1)

# Output the stylized dataframe using kable
kable(stylized_data, caption = "Stylized Random Data")

```




<br/><br/>


# 7. Demonstration of the `stylize2()` function

`stylize2()` functions similarly to `stylize1()`, but it aims to round values to 2 decimals instead.


```{r, echo=TRUE}

stylized_data <- random_df %>%
  mutate_at(vars("Group A", "Group B", "Group C"), stylize2)

# Output the stylized dataframe using kable
kable(stylized_data, caption = "Stylized Random Data")

```

<br/><br/>



# 8. Demonstration of the `stylize_p()` function

The `stylize_p()` function is designed to round and format numeric values, particularly for improving readability in statistical outputs. It follows specific rules to achieve this:

- Values are rounded to 2 decimal places whenever feasible.
- The function specifically avoids outputting "0.05", as this value may not hold practical meaning for the reader when the significance level (alpha) is set at 0.05.
- If the rounded result is equal to "0.050", the original value is retained without any rounding.
- If the original value (prior to rounding) is equal to 0.050, the original value is retained without any rounding.
- Values less than 0.001 are expressed as "<0.001".

<br/>

First, let's examine the regression results from an imaginary analysis.

<br/>

```{r, echo=FALSE}

library(tibble)

# Creating an updated imaginary tibble with more data
random_regression_results <- tibble(
  Outcome = c("Variable 1", "Variable 3", "Variable 5", "Variable 9"),
  Regressor = c("Variable 2", "Variable 4", "Variable 6", "Variable 10"),
  Beta = c(0.1234567, 0.9876543, 0.54321, 0.3456),
  SE = c(0.0123, 0.0456, 0.00789, 0.0543),
  `P-value` = c(0.001234, 0.056789, 0.000987, 0.0499)
)
 
# Displaying the data frame
kable(random_regression_results, caption = "random_regression_results")

```

<br/>

We can now apply `stylize2()` for the beta and SE columns and `stylize_p()` for the P-values.




```{r, echo=TRUE}

stylized_data <- random_regression_results %>%
  mutate_at(vars("Beta", "SE"), stylize2)

stylized_data <- stylized_data %>%
  mutate_at(vars("P-value"), stylize_p)

# Output the stylized dataframe using kable
kable(stylized_data, caption = "Stylized Regression Results")

```

<br/>
<br/>
<br/>
<br/>
<br/>
<br/>





  
  
  
